Reassessing the role of theory and machine learning in forecasting civil conflict

### 2020-09-21 paper.Rmd
work on discussion and conclusion
still in progress

@@

However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intended to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

Most social science research uses a weak form of theory.  In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.  Operationally, this boils down to four types of concepts: demands, accommodations,
nonviolent repression, and low-level violence.
\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}
These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article. The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict.  The actual mechanisms of the dynamism among these variables are not specified.

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
At this level of detail the theoretical model is more about chosing the correct variables.  And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.

@@

>>>>

# Discussion

However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:

1. A theory is separate from the case or cases being examined. It is intended to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

Most social science research uses a weak form of theory.  In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.  Operationally, this boils down to four types of concepts: demands, accommodations,
nonviolent repression, and low-level violence.

\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}

These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article. The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict.  The actual mechanisms of the dynamism among these variables are not specified.

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

At this level of detail the theoretical model is more about choosing the correct variables, and this is the contrast between theoretical prediction models and a-theoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" process at step 1.5 above, with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.

And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.

Another way in which we could contrast theoretical from a-theoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Probably it's more complex, and probably we don't really know what a better specification would be. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are whatever those relationships are linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

WHAT CAN ML MODELS TELL US ABOUT THEORY? PRUNING VARIABLES, FUNCTIONAL FORM, INTERACTIONS, PREDICABILITY

@@

@@

More importantly, trying to contrast machine learning and theory is misguided. Even if a "theory" section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models. But we do not. What we instead have are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our theories.
